---
title: "MIC regression"
date: "2025-11-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkgs, message=FALSE}
library(tidyverse)
library(cowplot)
library(readxl)
library(viridisLite)

# install_github("lu-harr/iddoPal")
library(iddoPal)
# install_github("lu-harr/looseVis")
library(looseVis)

dat <- read_excel("GAS_SR_Raw_50_90_NOS_formatted_20251106.xlsx")
```

Here are some quick summary stats:

```{r}
# number of isolates
sum(dat$count)

# number of isolates per country
dat %>%
  group_by(country) %>%
  summarise(n = sum(count)) %>%
  arrange(desc(n))

# number of isolates per who region
dat %>%
  group_by(who) %>%
  summarise(n = sum(count)) %>%
  arrange(desc(n))

# combined mic
mic <- dat %>%
  mutate(nos_90_comb = ifelse(is.na(micnos_1), mic90_1, micnos_1))

which(is.na(mic$nos_90_comb))

nos_vec <- rep(mic$micnos_1, mic$count)
median(nos_vec, na.rm=TRUE)
mean(nos_vec, na.rm=TRUE)
hist(nos_vec, breaks=100)
abline(v = median(nos_vec, na.rm=TRUE), col="blue")
abline(v = mean(nos_vec, na.rm=TRUE), col="red")

# represent entries with multiple isolates multiple times:
nos_90_comb_vec <- rep(mic$nos_90_comb, mic$count)
length(nos_90_comb_vec)
median(nos_90_comb_vec)
quantile(nos_90_comb_vec, probs = c(0.1, 0.5, 0.9))
range(nos_90_comb_vec)
summary(nos_90_comb_vec)
hist(nos_90_comb_vec, breaks = 100)
```

And now onto the regression ...

#### Populate names of WHO regions:

```{r who}
who_df <- matrix(c(c("1", "Africa"),
                c("2","Americas"),
                c("3", "Eastern Mediterranean"),
                c("4","Europe"),
                c("5", "Southeast Asia"),
                c("6", "Western Pacific")),
                ncol = 2, byrow = TRUE) %>%
  data.frame() %>%
  setNames(c("nume", "name"))

who_lst = who_df$name
names(who_lst) = who_df$nume
  
# and join to data:
dat <- dat %>%
  mutate(who = as.factor(who)) %>%
  left_join(who_df, by = join_by(who == nume)) %>%
  mutate(who_name = ifelse(is.na(name), "Not assigned", name)) %>%
  # need to put both MIC_nos and MIC_90 in one column
  mutate(mic_mix = case_when(!is.na(micnos_1) ~ micnos_1,
                             !is.na(mic90_1) ~ mic90_1,
                             TRUE ~ NA),
         weight_mic = 1 / mic_mix)

```

#### Variance-weighted least squares

```{r}
mod_unweighted = lm(mic_mix ~ isolate_yr, dat)
summary(mod_unweighted)

# setting weight to mean absolute residuals
wts <- data.frame(isolate_yr = dat$isolate_yr,
                  resid = mod_unweighted$residuals) %>%
  group_by(isolate_yr) %>%
  summarise(wt = median(abs(resid)))

dat <- dat %>%
  left_join(wts)

# a quick summary:
par(mfrow = c(1,2))
pred <- predict(mod_unweighted, dat)
plot(dat$isolate_yr, dat$mic_mix, cex = scale_cex(dat$count, max_cex = 5), main = "(unweighted) ordinary least squares")
abline(mod_unweighted, col="red")
plot(pred, dat$mic_mix - pred, main = "Residuals over fitted")

# now apply weights to mic_mix
```


#### Here's the bubble plot with both the MIC_nos and the MIC_90s:

```{r}
yticks <- c(0, 0.004, 0.008, 0.016, 0.032, 0.064, 0.12)

p <- ggplot(dat %>%
         filter(!is.na(mic_mix)) %>%
         # put smaller points in front:
         arrange(desc(count)),
       aes(x = isolate_yr, y = mic_mix)) +
  geom_point(aes(col = who_name, 
                 size = count), alpha = 0.5) +
  # pop a line on there
  # geom_smooth(method = "lm", se = FALSE, col = "darkgrey") +
  scale_color_discrete(name = "WHO Region") +
  scale_size_continuous(range = c(1,10), name = "Number of isolates") +
  # set specific ticks on y axis: (could transform this too if we think we need to ...)
  scale_y_continuous(breaks = yticks, minor_breaks = NULL, name = "MIC_nos") +
  labs(title = "MIC_nos + MIC_90 over time (points not grouped)") +
  theme_bw()

p
```

#### Now here's a bubble plot with regression:

```{r}
gap <- 1

# cobble something together ...
pal <- c(rev(viridis(4)),  
         pals$loosepal[3],
         iddo_palettes$iddo[1],
         "orange") %>% 
  unname()

# model weighted with weights from above
mod_weighted <- lm(mic_mix ~ isolate_yr, dat, weights = wt)
summary(mod_weighted)
confint(mod_weighted)
tmp <- predict(mod_weighted, data.frame(isolate_yr = 1928:2017), se.fit = TRUE)

p <- ggplot(dat %>%
            filter(!is.na(mic_mix)) %>%
            group_by(isolate_yr, mic_mix, who_name) %>%
            summarise(count = sum(count)) %>%
            mutate(who_name = factor(who_name, levels = c("Africa",
                                                          "Americas",
                                                          "Eastern Mediterranean",
                                                          "Europe",
                                                          "Southeast Asia",
                                                          "Western Pacific",
                                                          "Not assigned"))) %>%
         # put smaller points in front:
         arrange(desc(count)),
       aes(x = isolate_yr, y = mic_mix)) +
  coord_cartesian(ylim = c(0, 0.125)) +
  # pop a line on there
  geom_smooth(data = dat %>% filter(!is.na(mic_mix)),
              #mapping = aes(weight = count),
              #mapping = aes(weight = weight_mic),
              mapping = aes(weight = wt),
              method = "lm", se = TRUE, col = "darkgrey", linewidth = 0.8) +
  #geom_smooth() +
  geom_point(aes(col = who_name, 
                 size = count), alpha = 0.5) +
  scale_color_discrete(name = "WHO Region", type = pal) +
  # scale_size_continuous(range = c(1,6), name = "Number of isolates",
  #                       trans="log10") +
  scale_size_continuous(range = c(1,10), name = "Number of isolates") +
  # set specific ticks on y axis: (could transform this too if we think we need to ...)
  scale_y_continuous(breaks = yticks, minor_breaks = NULL, name = "MIC"#,
                     #limits = c(0, 0.125)
                     ) +
  scale_x_continuous(limits = c(1928 - gap, 2017 + gap)) +
  #labs(title = "MIC_nos + MIC_90 over time (points grouped, variance-weighted LS)") +
  labs(x = "Year", ylab = "MIC") +
  theme_bw()

p

ggsave("figs/mic_mixed_bubbles.png", 
       height = 6, width = 9, scale = 0.8)

```

#### Here's the same plot with some breaks in the x axis:

```{r}
# would love to transform the y axis but if i can't extract the ribbon from
# geom_smooth then I can't tell it not to freak out at zero ..

# sad - doesn't look like coord_cartesian will work with axis breaks
# https://github.com/YuLab-SMU/ggbreak/issues/59
library(ggbreak)
p + scale_x_break(c(1928 + gap, 1950 - gap)) +
  coord_cartesian(ylim = c(0, 0.125)) +
  scale_x_break(c(1950 + gap, 1968)) +
  scale_x_continuous(breaks = c(1928, 1950, seq(1960, 2021, 10)), 
                     limits = c(1928 - gap, 2016 + gap)) +
  geom_hline(aes(yintercept = 0)) +
  xlab("Year") + ylab("MIC")

ggsave("figs/mic_mixed_bubbles_with_breaks.png", 
       height = 6, width = 9, scale = 0.8)
# make colours viridis plus an orange maybe?
```

#### Now here's a plot with the y-axis logged but *not* the regression: 
(so regression's still on a linear scale)

```{r}
Y_LOWER <- 0.0018
X_LOWER <- 1925
mod_weighted <- lm(mic_mix ~ isolate_yr, dat, weights = wt)
confint(mod_weighted)

ci <- predict(lm(mic_mix~isolate_yr, dat, weights = wt), 
                 newdata = data.frame(isolate_yr = seq(X_LOWER,
                                                       max(dat$isolate_yr) + 2)),
                 interval = "confidence") %>%
  as.data.frame() %>%
  mutate(lwr = ifelse(lwr < Y_LOWER, Y_LOWER, lwr),
         isolate_yr = seq(X_LOWER, max(dat$isolate_yr) + 2))

# okay: need to transform y-axis and truncate
gap <- 1

# cobble something together ...
pal <- c(rev(viridis(4)),  
         pals$loosepal[3],
         iddo_palettes$iddo[1],
         "orange") %>% 
  unname()

yticks <- c(0, 0.002, 0.004, 0.008, 0.016, 0.032, 0.064, 0.12)

library(ggbreak)
p <- ggplot(dat %>%
            filter(!is.na(mic_mix)) %>%
            group_by(isolate_yr, mic_mix, who_name) %>%
            summarise(count = sum(count)) %>%
            mutate(who_name = factor(who_name, levels = c("Africa",
                                                          "Americas",
                                                          "Eastern Mediterranean",
                                                          "Europe",
                                                          "Southeast Asia",
                                                          "Western Pacific",
                                                          "Not assigned"))) %>%
         arrange(desc(count))) +
  geom_ribbon(data = ci, aes(ymin = lwr, ymax = upr, x = isolate_yr),
              alpha = 0.5, fill = "grey", colour = NA) +
  geom_line(data = ci, aes(x = isolate_yr, y = fit), 
            colour = "darkgrey", linewidth = 0.8) +
  geom_point(aes(x = isolate_yr, y = mic_mix,
                 col = who_name, 
                 size = count), alpha = 0.5) +
  scale_color_discrete(name = "WHO Region", type = pal) +
  scale_size_continuous(range = c(1,10), name = "Number of isolates") +
  scale_y_continuous(breaks = yticks, minor_breaks = NULL, name = "MIC") +
  scale_x_continuous(limits = c(X_LOWER, 2017 + 2)) +
  labs(x = "Year", ylab = "MIC") +
  coord_trans(y = "log2", ylim = c(Y_LOWER, 0.14), expand = FALSE) +
  theme_bw()

p

ggsave("figs/mic_mixed_bubbles_logged.png", 
       height = 6, width = 9, scale = 0.8)

```

The bounds of the confidence interval for mean MIC get a bit weird at low MICs (early years). What scale are we interpreting change in mean MIC on? Probably on the log scale also ... for this reason, we do the regression on the log scale below.

#### Alternatively, do the regression on the log scale and re-exp the labels

This is the model + visualisation we include in the manuscript. Given tested MICs are doubling, we carry out regression on the log2 scale.

```{r}
# apply log to the mics:
dat <- dat %>%
  mutate(mic_logged = log2(mic_mix))

# fit an unweighted model to determine weights
mod_unweighted = lm(mic_logged ~ isolate_yr, dat)
summary(mod_unweighted)

# setting weight to mean absolute residuals
wts <- data.frame(isolate_yr = dat$isolate_yr,
                  resid = mod_unweighted$residuals) %>%
  group_by(isolate_yr) %>%
  summarise(wt = median(abs(resid)))

dat <- dat %>%
  left_join(wts)

mod_weighted <- lm(mic_logged ~ isolate_yr, dat, weights = wt)
confint(mod_weighted)
summary(mod_weighted)

# this is slope on log-scale:
coefficients(mod_weighted)[2]
2**(coefficients(mod_weighted)[2])
2**confint(mod_weighted)[2,]

# Here's what that looks like unlogged:
plot(dat$isolate_yr, dat$mic_mix)
yyy = seq(min(dat$isolate_yr), max(dat$isolate_yr))
lines(yyy, 2**predict(mod_weighted, data.frame(isolate_yr = yyy)))
# (pretty flat)

Y_LOWER <- log2(0.0018)
X_LOWER <- 1925

ci <- predict(lm(mic_logged ~ isolate_yr, dat, weights = wt), 
                 newdata = data.frame(isolate_yr = seq(X_LOWER,
                                                       max(dat$isolate_yr) + 2)),
                 interval = "confidence") %>%
  as.data.frame() %>%
  mutate(lwr = ifelse(lwr < Y_LOWER, Y_LOWER, lwr),
         isolate_yr = seq(X_LOWER, max(dat$isolate_yr) + 2))

gap <- 1


# cobble something together ...
pal <- c(rev(viridis(4)),  
         pals$loosepal[3],
         iddo_palettes$iddo[1],
         "orange") %>% 
  unname()

yticks <- c(0, 0.002, 0.004, 0.008, 0.016, 0.032, 0.064, 0.12)

p <- ggplot(dat %>%
            filter(!is.na(mic_mix)) %>%
            group_by(isolate_yr, mic_mix, who_name) %>%
            summarise(count = sum(count)) %>%
            mutate(who_name = factor(who_name, levels = c("Africa",
                                                          "Americas",
                                                          "Eastern Mediterranean",
                                                          "Europe",
                                                          "Southeast Asia",
                                                          "Western Pacific",
                                                          "Not assigned"))) %>%
         arrange(desc(count))) +
  geom_ribbon(data = ci, aes(ymin = 2**lwr, ymax = 2**upr, x = isolate_yr),
              alpha = 0.5, fill = "grey", colour = NA) +
  geom_line(data = ci, aes(x = isolate_yr, y = 2**fit), 
            colour = "darkgrey", linewidth = 0.8) +
  geom_point(aes(x = isolate_yr, y = mic_mix,
                 col = who_name, 
                 size = count), alpha = 0.5) +
  scale_color_discrete(name = "WHO Region", type = pal) +
  scale_size_continuous(range = c(1,10), name = "Number of isolates", breaks = c(1, 10, 100, 1000)) +
  scale_y_continuous(breaks = yticks, minor_breaks = NULL, name = "MIC") +
  scale_x_continuous(limits = c(X_LOWER, 2017 + 2), expand = c(0,0)) +
  labs(x = "Year", ylab = "MIC") +
  coord_trans(y = "log2") + #, ylim = c(Y_LOWER, 0.14), expand = FALSE) +
  theme_bw()

p

ggsave("figs/mic_mixed_bubbles_all_logged.png", 
       height = 6, width = 9, scale = 0.8)

plot(mod_weighted)
```

#### Ordinal regression

We could really treat MICs as categories as intermediate MICs are not (usually) tested. For this reason, we've had a look at ordinal regression in the chunk below.


```{r}
library(MASS)

ftable(dat$mic_mix)

# say we bin MIC ..?
bins <- c(0.002, 0.004, 0.008, 0.016, 0.032, 0.064, 0.12)
dat$mic_mix_binned <- cut(dat$mic_mix,
                          breaks = c(0, bins, max(dat$mic_mix)),
                          ordered_result = TRUE)
# I've totally forgotten how we got to MIC90s and MICNosses so this ^ may not be 
# appropriate

ftable(dat$mic_mix_binned)

# Fit the model - no need to log: that would be for visualisation purposes only under ordinal model
# tried using wts from logged model ......... not thinking too hard about that but we should at some point
# it looks like weights need to be integers as we end up assuming binomial glm for ordinal *logistic* regression
# haven't thought too hard about this ... may need to be probit link
model <- polr(mic_mix_binned ~ isolate_yr, data = dat, Hess = TRUE)

summary(model)
# note the slope here

new_data = data.frame(isolate_yr = seq(X_LOWER, max(dat$isolate_yr) + 2))
preds <- predict(model, 
                 newdata = new_data,
                 type = "probs") %>%
  as.data.frame() %>%
  mutate(max_probs = max.col(across(starts_with("(")))) %>%
  mutate(isolate_yr = new_data$isolate_yr)

preds_long <- pivot_longer(preds, starts_with("(")) %>%
  mutate(lower = str_extract(name, "(?<=\\()[^,]+") %>% as.numeric(),
         upper = str_extract(name, "(?<=,)[^]]+") %>% as.numeric(),
         height = lower + (upper - lower)*value) %>%
  filter(lower != 0)

# so now we have colour bars representing class probability between the lower 
# and upper extents of the bins ... which don't change much because there isn't
# much change in MIC slope over time
ggplot(data = dat) +
  geom_point(aes(x = isolate_yr, y = mic_mix)) +
  scale_y_continuous(trans = "log2", breaks = bins) +
  geom_ribbon(data = preds_long, aes(ymin = lower, ymax = height, fill = name, x = isolate_yr), 
              alpha = 0.5) +
  scale_fill_discrete("MIC bin") +
  # and this black line is the predicted class, which stays the same for the duration of the data
  # (the -1 is because I removed the bottom bin)
  geom_hline(yintercept = bins[preds$max_probs[1] - 1])

```

In the end, we didn't go with ordinal regression, as the class with the greatest probability did not change over time! And the linear regression gave us a similar result!


